The Categorical Structure of Profunctor Optics

### 2017-07-26 paper.tex
Start filling out Prisms and Setters

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\PreOptic}{\mathbf{PreOptic}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Tamb}{\mathbf{Tamb}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}

@@

\todo{Might have to adjust the entire thing because the action of $\M$ on $\C$ is not strictly monoidal. On the other hand, through the entire Pastro/Street paper pretends associativity/etc is strict, so maybe who cares? What we really want is just a (non-strict) monoidal functor $\M \to [\C, \C]$ from some $\M$.}

@@

As remarked in the introduction \todo{(todo: remark in the introduction)}, the preoptics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\PreOptic_\M((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

\subsubsection{Prisms}
\subsubsection{Traversals}
\subsubsection{Setters}

@@

\subsection{Tambara Modules}
The following section generalises definitions that first appeared in \cite{Doubles}.
\begin{definition}
Suppose $\C$ is acted on by $\M$, a subcategory of $[\C, \C]$ closed under composition, and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a natural family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
such that $\alpha_{A,B,\id_\C} = \id_{P(A,B)}$, and the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, MN}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
\end{definition}

@@

\section{Degenerate Optics}
% \section{Optic Families}
\section{Indexed Optics}
\todo{
This possibly looks like
\begin{align*}
\int^{M \in \M} \C(S, M(IA)) \times \C(M A', S')
\end{align*}
with another monoidal action $I$. The $I$ would need to distribute over $M$ in some way.
}
\section{Merely Well-Behaved Optics}
\todo{
The lenses that the UPenn crew look at are less restricted. They obey the $\fput\fget$ and $\fget\fput$ laws but not the $\fput\fput$ law. 
}

@@

\begin{example}
A minimal example is the following:
\begin{align*}
\fget : \{A,B,C\} &\to \{X, Y\} \\
A &\mapsto X \\
B &\mapsto X \\
C &\mapsto Y 
\end{align*}
\begin{align*}
\fput : \{A,B,C\} \times \{X, Y\} &\to \{A,B,C\} \\
(A,X) &\mapsto A \\
(B,X) &\mapsto B \\
(C,X) &\mapsto B \\
(A,Y) &\mapsto C \\
(B,Y) &\mapsto C \\
(C,Y) &\mapsto C
\end{align*}
\end{example}

\todo{THE FOLLOWING IS TOTALLY WRONG, does not capture the right thing, fails for the above example}

@@

\begin{proposition}
The converse is true.
\end{proposition}
\begin{proof}

\end{proof}

This suggests the following definition of merely well-behaved optics, which we \todo{may as well} call semi-optics.

>>>>

\newcommand{\Act}{\mathbf{Act}}
\newcommand{\PreOptic}{\mathbf{PreOptic}}
\newcommand{\Optic}{\mathbf{Optic}}
\newcommand{\SemiOptic}{\mathbf{SemiOptic}}
\newcommand{\Lens}{\mathbf{Lens}}
\newcommand{\Prism}{\mathbf{Prism}}
\newcommand{\Hask}{\mathbf{Hask}}
\newcommand{\Endo}{\mathbf{Endo}}
\newcommand{\Strong}{\mathbf{Strong}}
\newcommand{\Tamb}{\mathbf{Tamb}}
\newcommand{\Point}{\mathbf{Point}}
\newcommand{\CoPoint}{\mathbf{CoPoint}}
\newcommand{\Traversable}{\mathbf{Traversable}}

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}
\newcommand{\inl}{\mathrm{inl}}
\newcommand{\inr}{\mathrm{inr}}

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}

@@

\todo{TODO: Might have to adjust the entire thing because the action of $\M$ on $\C$ is not strictly monoidal. On the other hand, through the entire Pastro/Street paper pretends associativity/etc is strict, so maybe who cares? What we really want is just a (non-strict) monoidal functor $\M \to [\C, \C]$ from some $\M$.}

\todo{TODO: Think about size issues, we are taking some very big coends...}

@@

As remarked in the introduction \todo{(todo: remark in the introduction)}, the preoptics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\PreOptic_\times((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

\subsubsection{Prisms}
Prisms are dual to lenses:

\begin{definition}
Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} in $\C$ is the category of optics with respect to this action: $\Prism = \Optic_\sqcup$.
\end{definition}

Just as lenses are given by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = [r\inl, \id_S]l$

The laws for prisms are:
\begin{align*}
\fmatching \; \freview = \inr \\
[\id_S, \freview] \fmatching = \id_S
\end{align*}

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
If $\langle l, r \rangle : S \hto A$ is a prism then the associated $\fmatching$ and $\freview$ functions satisfy the above laws.
\end{proposition}
\begin{proof}
\todo{Should be straightforward}
\end{proof}

\todo{unfinished}

\subsubsection{Traversals}
\subsubsection{Setters}

To define setters we must first recall the notion of a strong functor.

\todo{Taken from http://cs.ioc.ee/~tarmo/ssgep15/ssgep-1a.pdf}
\begin{definition}
\todo{Again I am not fussing with associators/unitors, this is probably bad. Also this definition should work fine of we instead use an action of some $\M$ on $\C$.} Suppose $(\C, \otimes, I)$ is a monoidal category. A \emph{(left) strength} for an endofunctor $F : \C \to \C$ is a natural transformation
\begin{align*}
\theta_{A,B} : A \otimes F B \to F(A \otimes B)
\end{align*}
such that both
\[
\begin{tikzcd}
A \otimes B \otimes F C \ar[r, "A \otimes \theta_{B,C}"] \ar[d, "\theta_{A \otimes B, C}" left]  & A \otimes F(B\otimes C) \ar[d, "\theta_{A, B\otimes C}" right] \\
F(A \otimes B \otimes C) \ar[r, equals]& F(A \otimes B \otimes C)
\end{tikzcd}
\]
and
\[
\begin{tikzcd}
1 \otimes F A \ar[r, "s_{1,A}"] \ar[d, "\cong" left]  & F(1 \otimes A) \ar[d, "\cong" right] \\
F A \ar[r, equals] & F A
\end{tikzcd}
\]
commute. A \emph{strong functor} is a functor equipped with a strength. A \emph{strong natural transformation} $\tau : (F,\theta) \to (G,\theta')$ is a natural transformation that respects the strength:
\[
\begin{tikzcd}
A \otimes F B \ar[r, "\theta_{A,B}"] \ar[d, "A \otimes \tau_B" left]  & G(A \otimes B) \ar[d, "\tau_{A \otimes B}" right] \\
A \otimes G B \ar[r, "\theta'_{A, B}"]& G(A \otimes B)
\end{tikzcd}
\]

There is an evident category $\Strong(\C)$ of strong endofunctors and strong natural transformations, and a forgetful functor $U : \Strong(\C) \to [\C, \C]$.
\end{definition}

Every endofunctor on $(\Set, \times, 1)$ has a unique strength, so this concept is invisible for Haskell functors.

\begin{lemma}
Suppose $\C$ is cartesian closed and $F : \C \to \C$ is strong with respect to $\times$. Then for any objects $A, B \in \C$, \[\C(A, FB) \cong \Strong_\C(\times)(A \times \homC(B, -), F)\]
\end{lemma}
\begin{proof}
Given a $f \in \C(A, FB)$, let $\eta$ be the natural transformation with components
\begin{align*}
A \times \homC(B, C) \xrightarrow{f \times \id} FB \times \homC(B, C) \xrightarrow{\theta_{\homC(B, C), B}}  F(B \times \homC(B, C)) \xrightarrow{F\epsilon} FC
\end{align*}
where $\theta$ is the strength for $F$ and $\epsilon$ is the evaluation map.

Conversely, given an $\eta : A \times \homC(B, -) \to F$, the component at $B$ is a function $g : A \times \homC(B, B) \to FB$; we recover $f : A \to FB$ as $f = g (\id_A, \const_{\id_B})$.

\todo{These are inverses.}
\end{proof}

This is a generalisation of \cite[Proposition 2.2]{SecondOrderFunctionals}

\todo{unfinished}

\subsubsection{Affine Traversals}

@@

\subsection{Tambara Modules}
The following section generalises definitions that first appeared in \cite{Doubles}.
\begin{definition}
Suppose $\C$ is acted on by $(\M, \otimes, I)$ and let $P \in \Prof(\C, \C)$ be a profunctor. A \emph{Tambara module structure for $\M$} on $P$ is a natural family of maps:
\begin{align*}
\alpha_{A,B,M} : P(A,B) \to P(MA, MB)
\end{align*}
such that $\alpha_{A,B,I} = \id_{P(A,B)}$, and the following diagram commutes:
\[
\begin{tikzcd}
P(A,B) \ar[r, "\alpha_{A,B,M}"] \ar[dr, "\alpha_{A, B, N \otimes M}" below left] & P(MA, MB) \ar[d, "\alpha_{MA, MB, N}" right] \\
& P(NMA, NMB)
\end{tikzcd}
\]
\end{definition}

@@

\section{Degenerate Optics}
\todo{The Haskell lens library has some other `optics' that aren't proper optics, but still integrate nicely with the rest of the ecosystem, e.g. Getters, Reviews and Folds}

\section{Optic Families}
\section{Indexed Optics}
\todo{
This possibly looks like
\begin{align*}
\int^{M \in \M} \C(S, M(IA)) \times \C(M A', S')
\end{align*}
with another monoidal action $I$. The $I$ would need to distribute over $M$ in some way. ``Coindexed'' optics seem to have 
\begin{align*}
\int^{M \in \M} \C(S, J(MA)) \times \C(M A', S')
\end{align*}
which is why they pass information `in the other direction' but still transform covariantly.
}
\section{Merely Well-Behaved Optics}
\todo{
The lenses that the UPenn crew look at are less restricted. They obey the $\fput\fget$ and $\fget\fput$ laws but not the $\fput\fput$ law. 
}

@@

\begin{example}
A minimal example is the following:
\begin{align*}
\fget : \{A,B,C\} &\to \{X, Y\} \\
A &\mapsto X \\
B &\mapsto X \\
C &\mapsto Y 
\end{align*}
\begin{align*}
\fput : \{A,B,C\} \times \{X, Y\} &\to \{A,B,C\} \\
(A,X) &\mapsto A \\
(B,X) &\mapsto B \\
(C,X) &\mapsto B \\
(A,Y) &\mapsto C \\
(B,Y) &\mapsto C \\
(C,Y) &\mapsto C
\end{align*}
\todo{So here the $\fput\fget$ composite is
\begin{align*}
(A,X) &\mapsto (A,X) \\
(B,X) &\mapsto (B,X) \\
(C,X) &\mapsto (B,X) \\
(A,Y) &\mapsto (C,Y) \\
(B,Y) &\mapsto (C,Y) \\
(C,Y) &\mapsto (C,Y)
\end{align*}
}
\end{example}

\todo{The condition on $\langle l, r \rangle$ seems to be that $rl = \id_S$ and $\pi_2lr = \pi_2$. I can't see how to generalise this second condition to optics other than lenses.}

\todo{The following original attempt is wrong, does not capture the right thing}

@@

This suggests the following definition of merely well-behaved optics, which we may as well call semi-optics.
